Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-03 submission.tex
Experimental setup nearly finished

\cite{Meg:01,Meg:02} trained and evaluated a range of PoS taggers on the
Stockholm-Umeå Corpus (SUC) \cite{Gus:Har:06}, annotated with a Swedish version
of PAROLE tags, with the tag set totaling 139 tags. Furthermore, they
investigated the effects of tag set size on tagging by mapping the original tag
set into smaller subsets designed for parsing. They argue that a tag set with
complete morphological tags may not be necessary for all NLP applications, for
instance syntactic parsing. They found that the smallest tag set comprising 26
tags yields the lowest tagger error rate. However, for some of the taggers,
augmenting the tag set with more linguistically informative tags may actually
lead to a drop in error rate. They argue that this shows that the size of the
tag set as well as the type of information in the tags are crucial factors for
tagger performance. However/unfortunately, they do not report results of parsing
with the various PoS tag sets.

Similarly, \newcite{Mac:05} investigated the effects of PoS tag sets on tagger
performance in English, specifically the Wall Street Journal portion of Penn
Treebank \cite{Mar:San:Mar:93}. They mapped the original tag set of Penn
Treebank to more fine-grained tag sets using linguistic insight to investigate
whether the additional linguistic information included in the finer-grained
tags could assist the tagger. Experimenting with both lexically and
syntactically conditioned modifications such as distinguishing between count
nouns and noncount nouns as well as between transitive and intransitive verbs,
they found that more fine-grained tag sets rarely led to improvements in tagger
accuracy; the most successful modification yielded an improvement in tagger
accuracy of 0.05 percentage points.

@@

\section{Data}
\label{sec:data}
\subsection{The Norwegian Dependency Treebank}
We used the newly developed Norwegian Dependency Treebank (NDT)
\cite{Sol:Skj:Ovr:14}, which is the first publicly available treebank for
Norwegian. It was developed at the National Library of Norway in collaboration
with the University of Oslo, and contains manual syntactic and morphological
annotation. The treebank contains 311 000 tokens of Bokmål and 303 000 tokens
of Nynorsk.  The annotated texts are mostly newspaper text, but also include
government reports, parliament transcripts and excerpts from blogs. The
annotation process of the treebank was supported by the Oslo-Bergen Tagger
\cite{Hag:Joh:Nok:00} and then manually corrected by annotators.

@@

\paragraph{Morphological Annotation}
The morphological annotation and PoS tag set of NDT follows the Oslo-Bergen
Tagger \cite{Hag:Joh:Nok:00,Sol:13}, which in turn is largely based on the work
of \newcite{Faa:Lie:Van:97}. The tag set consists of 12 morphosyntactic PoS
tags, with 7 additional tags for punctuation and symbols.  The tag set is thus
rather coarse-grained, with broad categories such as \texttt{subst} (noun) and
\texttt{verb} (verb). The PoS tags are complemented by a large set of
morphological features, providing information about morphological properties such as
definiteness, number and tense. These features are used in our tag set
modifications, where the coarse PoS tag of relevant tokens are concatenated
with one or more of these features.  This allows us to include more linguistic
information in the tags.

@@

\paragraph{Syntactic Annotation}
The syntactic annotation choices in NDT are largely based on the Norwegian
Reference Grammar \cite{Faa:Lie:Van:97}. The annotation choices are outlined in
Table \ref{ndtannotation}, taken from \newcite{Sol:Skj:Ovr:14}, providing
overview of the analyses of syntactic constructions that often distinguish
dependency treebanks, such as coordination and the treatment of auxiliary and
main verbs. The set of dependency relations contains 29 dependency relations,
including \textsc{adv} (adverbial), \textsc{subj} (subject) and \textsc{koord}
(coordination).

@@

\paragraph{Data Set Split}
As there was no standardized data set split of NDT due to its very recent
development, we needed to establish a data set split
(training/development/test) in preparation to our experiments. Our data set
split of the treebank follows the standard 80-10-10 (training/development/test)
split and will be distributed with the treebank and proposed as the new
standard(?). In creating the data set split, care has been taken to preserve
contiguous texts in the various data sets while keeping the split balanced in
terms of genre (and source). Our proposed data set split was used in the
Norwegian contribution to the Universal Dependencies project \cite{Ovr:Hoh:16}.
The split will be made available at a companion website.

\paragraph{Tagger}
For our experiments with tag set modifications, we wanted a PoS tagger that is
both fast and accurate. There is often a trade-off between the two, as the best
taggers tend to suffer in terms of speed due to their complexity. However, a
tagger that achieves both close to state-of-the-art accuracy as well as very
high speed is TnT \cite{Bra:00}. The fact that TnT was used for evaluating the
universal tag set \cite{Pet:Das:McD:12}, served as another good indication of
TnT being appropriate for our task. The sum of these factors led to TnT being
the tagger of choice for our experiments.

\paragraph{Parser}
In choosing a syntactic parser for our experiments, we considered previous work
on dependency parsing of Norwegian, specifically that of \cite{Sol:Skj:Ovr:14}.
They found the Mate parser \cite{Boh:10} to be the most successful parser for
the parsing of NDT. Furthermore, recent dependency parser comparisons
\cite{Cho:Tet:Ste:15} showed that Mate performed very well on parsing of the
English portion of the OntoNotes 5 corpus, beating a range of contemporary
state-of-the-art parsers.

\paragraph{Tag Set Mapping}
In order to alter the tag set of NDT, we created a mapping for carrying out the
tag set modifications. We created a mapping for carrying out the tag set
modifications that maps the relevant existing tags to new, more fine-grained
tags including more relevant morphological features for the applicable tokens.

@@

\paragraph{Tags \& Features}
As we seek to quantify the effects of PoS tagging in a realistic setting, we
want to run the parser on automatically assigned PoS tags. For the training of
the parser, however, we have two options: using either gold standard or
automatically assigned tags. In order to settle on a configuration, we
conducted experiments with gold standard and automatically assigned tags to see
how they differ with respect to performance. The results of our experiments
reveal that the combination of training and testing on automatic tags is
superior to training on gold standard tags and testing on automatic tags,
surprisingly. Consequently, the parser was both trained and tested on
automatically assigned tags in our experiments.

Note that it is absolutely crucial that the morphological features in the
treebank are removed when using automatic tags, as they are still gold
standard. For instance, if a verb token is erroneously tagged as a noun, we
could potentially have a noun token with verbal features such as tense, which
markedly obfuscates the training and parsing. Another important factor is that
we want to isolate the effect of PoS tags, necessitating the exclusion of
morphological features.

>>>>

\cite{Meg:01,Meg:02} trained and evaluated a range of PoS taggers on the
Stockholm-Umeå Corpus (SUC) \cite{Gus:Har:06}, annotated with a tag set based
on a Swedish version of PAROLE tags totaling 139 tags. Furthermore, they
investigated the effects of tag set size on tagging by mapping the original tag
set into smaller subsets designed for parsing. They argue that a tag set with
complete morphological tags may not be necessary for all NLP applications, for
instance syntactic parsing. They found that the smallest tag set comprising 26
tags yields the lowest tagger error rate. However, for some of the taggers,
augmenting the tag set with more linguistically informative tags may actually
lead to a drop in error rate. They argue that this shows that the size of the
tag set as well as the type of information in the tags are crucial factors for
tagger performance. However/unfortunately, they do not report results of
parsing with the various PoS tag sets.

Similarly, \newcite{Mac:05} investigated the effects of PoS tag sets on tagger
performance in English, specifically the Wall Street Journal portion of Penn
Treebank \cite{Mar:San:Mar:93}. They mapped the original tag set of Penn
Treebank to more fine-grained tag sets using linguistic insight to investigate
whether additional linguistic information included in finer-grained tags could
assist the tagger. Experimenting with both lexically and syntactically
conditioned modifications such as distinguishing between count nouns and
noncount nouns and between transitive and intransitive verbs, they found that
more fine-grained tag sets rarely led to improvements in tagger accuracy; the
most successful modification yielded an improvement in tagger accuracy of 0.05
percentage points.

@@

\section{Data}
\label{sec:data}
\subsection{The Norwegian Dependency Treebank}
We used the newly developed Norwegian Dependency Treebank (NDT)
\cite{Sol:Skj:Ovr:14},  the first publicly available treebank for Norwegian. It
was developed at the National Library of Norway in collaboration with the
University of Oslo, and contains manual syntactic and morphological annotation.
The treebank contains data from both varieties of Norwegian; 311 000 tokens of
Bokmål and 303 000 tokens of Nynorsk. The annotated texts are mostly newspaper
text, but also include government reports, parliament transcripts and excerpts
from blogs. The annotation process of the treebank was supported by the
Oslo-Bergen Tagger \cite{Hag:Joh:Nok:00} and then manually corrected by
annotators.

@@

\paragraph{Morphological Annotation}
The morphological annotation and PoS tag set of NDT follows the Oslo-Bergen
Tagger \cite{Hag:Joh:Nok:00,Sol:13}, which in turn is largely based on the work
of \newcite{Faa:Lie:Van:97}. The tag set consists of 12 morphosyntactic PoS
tags, with 7 additional tags for punctuation and symbols.  The tag set is thus
rather coarse-grained, with broad categories such as \texttt{subst} (noun) and
\texttt{verb} (verb). The PoS tags are complemented by a large set of
morphological features, providing information about morphological properties such as
definiteness, number and tense. These features are used in our tag set
modifications, where the coarse PoS tag of relevant tokens are concatenated
with one or more of these features to include more linguistic information in
the tags.

@@

\paragraph{Syntactic Annotation}
The syntactic annotation choices in NDT are largely based on the Norwegian
Reference Grammar \cite{Faa:Lie:Van:97}. The annotation choices are outlined in
Table \ref{ndtannotation}, taken from/courtesy of \newcite{Sol:Skj:Ovr:14},
providing overview of the analyses of syntactic constructions that often
distinguish dependency treebanks, such as coordination and the treatment of
auxiliary and main verbs. The set of dependency relations comprises 29
dependency relations, including \textsc{adv} (adverbial), \textsc{subj}
(subject) and \textsc{koord} (coordination).

@@

\paragraph{Data Set Split}
As there was no standardized data set split of NDT due to its very recent
development, we needed to establish a data set split
(training/development/test). Our data set split of the treebank follows the
standard 80-10-10 (training/development/test) split and will be distributed
with the treebank and proposed as the new standard(?). In creating the data set
split, care has been taken to preserve contiguous texts in the various data
sets while keeping the split balanced in terms of genre (and source). Our
proposed data set split was used in the Norwegian contribution to the Universal
Dependencies project \cite{Ovr:Hoh:16}.  The split will be made available at a
companion website.

\paragraph{Tagger}
For our experiments with tag set modifications, we wanted a PoS tagger both
reasonably fast and accurate. There is often a trade-off between the two
factors/considerations, as the most accurate taggers tend to suffer in terms of
speed due to their complexity. However, a tagger that achieves both close to
state-of-the-art accuracy as well as very high speed is TnT \cite{Bra:00}. The
fact that TnT was used for evaluating the universal tag set
\cite{Pet:Das:McD:12}, served as another good indication of TnT being
appropriate for our task. The sum of these factors led to TnT being the tagger
of choice for our experiments.

\paragraph{Parser}
In choosing a syntactic parser for our experiments, we considered previous work
on dependency parsing of Norwegian, specifically that of \cite{Sol:Skj:Ovr:14}.
They found the Mate parser \cite{Boh:10} to be the most successful parser for
the parsing of NDT. Furthermore, recent dependency parser comparisons
\cite{Cho:Tet:Ste:15} showed that Mate performed very well on parsing of
English, beating a range of contemporary state-of-the-art parsers.

\paragraph{Tag Set Mapping}
In order to carrying out the tag set modifications, we created a mapping that
maps the relevant existing tags to new, more fine-grained tags including more
relevant morphological features for the applicable tokens. A given tag set
modification involves specifying a tag and one or more features to be
concatenated with said tag for all applicable tokens, i.e., tokens that are
assigned said tag and contains the set of features in its morphological
features.

@@

\paragraph{Tags \& Features}
As we seek to quantify the effects of PoS tagging in a realistic setting, we
want to run the parser on automatically assigned PoS tags. For the training of
the parser, however, we have two options: using either gold standard or
automatically assigned tags. In order to settle on a configuration, we
conducted experiments with gold standard and automatically assigned tags to see
how they differ with respect to performance. The results of our experiments
reveal that the combination of training and testing on automatic tags is
superior to training on gold standard tags and testing on automatic tags.
Consequently, the parser was both trained and tested on automatically assigned
tags in our experiments.

We removed any morphological features in order to simulate a realistic setting.
Moreover, it is crucial that we remove these features when working with
automatically assigned tags, as the features is still gold standard.  For
instance, if a verb token is erroneously tagged as a noun, we could potentially
have a noun token with verbal features such as tense, which markedly obfuscates
the training and parsing. Another important aspect is that we want to isolate
the effect of PoS tags, necessitating the exclusion of morphological features.

%Note that it is absolutely crucial that the morphological features in the
%treebank are removed when using automatic tags, as they are still gold
%standard. For instance, if a verb token is erroneously tagged as a noun, we
%could potentially have a noun token with verbal features such as tense, which
%markedly obfuscates the training and parsing. Another important factor is that
%we want to isolate the effect of PoS tags, necessitating the exclusion of
%morphological features.
